---
title: "Package prep"
author: "Timothee Bonnet"
date: "22 June 2020"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Why create your own R-packages?

* Increases reproducibility
* Forces you to clean, check and document your code; which removes many mistakes and problems.
* Simple to do but looks so cool.
* Opens the door to putting packages along publications, Github, CRAN and be recognized and cited for your programming work.

## Packages you will need to create packages easily

Install if you do not have them already.

```{r}
library(devtools)
library(roxygen2)
```

> Note: there are several pipelines to create packages, some using directly RStudio tools. I chose to demonstrate using a pipeline that relies mostly on code and little on clicking buttons, because the former is more reproducible.

## Getting started

Start a R session from a parent directory in which you want to create your package. The package will be initialized inside a new directory. This new directory will also be a new RStudio project.

We will create a package that contains a few functions to handle temperature data. Our package will simply be called "temperature".

Let's initialize the package:
```{r, eval=FALSE}
create("temperature")
```

Check the content of the new directory "temperature":

* **A folder "R", currently empty. We will store our R-functions in there.**
* A file DESCRIPTION. Meta-data about your package.
* A file NAMESPACE. Handle relationships with other packages. Do not modify by hand. Not that important if you’re only developing packages for yourself; very important if you plan to share.
* An RStudio project file "temperature.Rproj". Don't worry about it.

Let's add some R-functions to the R-folder. Create a R-file called "functions.R" and containing the following two functions that form the base of our package:

```{r, eval=FALSE}
fahr_to_kelvin <- function(temp) {
  stopifnot(is.numeric(temp))
  kelvin <- ((temp - 32) * (5 / 9)) + 273.15
  return(kelvin)
}

kelvin_to_celsius <- function(temp) {
  celsius <- temp - 273.15
  return(celsius)
}

celsius_to_kelvin <- function(temp) {
  kelvin <- temp + 273.15
  return(kelvin)
}
```

You can create this file by hand or use some code:
```{r, eval=FALSE}
sink(file = "R/Functions.R")
cat("fahr_to_kelvin <- function(temp) {
  stopifnot(is.numeric(temp))
  kelvin <- ((temp - 32) * (5 / 9)) + 273.15
  return(kelvin)
}

kelvin_to_celsius <- function(temp) {
  celsius <- temp - 273.15
  return(celsius)
}

celsius_to_kelvin <- function(temp) {
  kelvin <- temp + 273.15
  return(kelvin)
}
")
sink()

```

Now we can load the functions and get ready to install your package:

```{r}
load_all(".")
```

```{r, eval=FALSE}
document(".")
```
For now `document()` only created the empty directory `man/` and nothing else. That is because there was nothing to document yet.

Now we can install the package from the parent directory:
```{r, eval=FALSE}
install("../temperature")
```

Your package already appears in R-Studio Package tab and can be used. However, there is not much information about the package and its functions.


Our functions are not accessible in R:
```{r, eval=FALSE}
fahr_to_kelvin(10)
```

And there is no information about them:
```{r, eval=FALSE}
?fahr_to_kelvin
```

Let's add more information to our package using Roxygen syntax. First for the `fahr_to_kelvin` function:

```{r, eval=FALSE}
#' Fahrenheit to Kelvin conversion
#' 
#' This function takes a temperature expressed in Fahrenheit and convert it to Kelvin.
#' @param temp A temperature expressed in Fahrenheit
#'
#' @export
fahr_to_kelvin <- function(temp) {
  stopifnot(is.numeric(temp))
  kelvin <- ((temp - 32) * (5 / 9)) + 273.15
  return(kelvin)
}


```

The `@export` line is critical. This tells Roxygen2 to add this function to the NAMESPACE file, so that it will be accessible to users. For your first R package, you’ll probably want to include `@export` for each of your functions (in large packages it is sometimes a good idea to hide some functions that are used only internally by other functions).

Now let's repeat the process:
```{r, eval=FALSE}
load_all(".")
document(".")
```
Notice that the NAMESPACE file has changed.

This time a file has been created in `/man`
```{r, eval=FALSE}
install("../temperature")
```

And we can reach our function help page and run the function:
```{r, eval=FALSE}
?fahr_to_kelvin
fahr_to_kelvin(10)
```

You can see that it is still missing some usual components that are for instance on the `mean()` help page: Value, References, See Also, Examples...

```{r, eval=FALSE}
#' Fahrenheit to Kelvin conversion
#' 
#' This function takes a temperature expressed in Fahrenheit and convert it to Kelvin.
#' @param temp A temperature expressed in Fahrenheit
#' 
#' @return A temperature in Kelvin
#' 
#' @examples
#' fahr_to_kelvin(32)
#' 
#' @export
fahr_to_kelvin <- function(temp) {
  stopifnot(is.numeric(temp))
  kelvin <- ((temp - 32) * (5 / 9)) + 273.15
  return(kelvin)
}

```

```{r, eval=FALSE}
load_all(".", reset = TRUE)
document(".")
install("../temperature", reload = TRUE)
```

```{r, eval=FALSE}
library(temperature)
?fahr_to_kelvin
fahr_to_kelvin(temp = )
```

By the way, sometimes you need to restart your R-session between installations for RStudio to be able to find help pages.

> Now create documentation for `kelvin_to_celsius()`, re-install the package and check you can access the new help page.

```{r, eval=FALSE}
#' Kelvin to Celsius conversion
#'
#' This function takes a temperature expressed in Kelvin and convert it to Celsius.
#'
#' @param temp A temperature expressed in Kelvin
#'
#' @return A temperature in Celsius.
#'
#' @examples
#' kelvin_to_celsius(300)
#'
#' @export
kelvin_to_celsius <- function(temp) {
  celsius <- temp - 273.15
  return(celsius)
}
```


### Initial fill of the DESCRIPTION

```{r, eval=FALSE}
Package: temperature
Title: Utilities to handle temperature data
Version: 0.0.1
Authors@R: 
    person(given = "Timothée",
           family = "Bonnet",
           role = c("aut", "cre"),
           email = "timotheebonnetc@gmail.com",
           comment = c(ORCID = "0000-0001-7186-5288"))
Description: What the package does (one paragraph).
License: GPL-3
Encoding: UTF-8
LazyData: true
Roxygen: list(markdown = TRUE)
RoxygenNote: 7.1.0
```


## More complete build

```{r, eval=FALSE}
#roxygenise(".", clean = TRUE)
load_all(".")
document(".")
build(pkg = ".", manual = TRUE)
check(".", manual = TRUE )
install("../temperature", reload = TRUE)

```

Check is very useful. Always try to get 0 errors, 0 warnings, 0 notes.

```{r}

```


More complete documentation:
```{r, eval=FALSE}
build_manual()
build_readme()
build_site()
```

At a later stage consider creating vignettes (i.e. tutorials, examples...)
```{r, eval=FALSE}
build_vignettes()
```



## Add data

Temperature data from Tuggeranong 1997-2019 http://www.bom.gov.au/climate/data/

```{r, eval=FALSE}
download.file("https://timotheenivalis.github.io/data/tuggeranong.csv", 
              destfile = "tuggeranong.csv")
tuggeranong <- read.csv("tuggeranong.csv")
```

Initialize the dataset
```{r, eval=FALSE}
use_data(tuggeranong, overwrite = TRUE)
```
We need to document the dataset. 
The documentation goes into the `/R` folder in a file called `data.R`.

```{r, eval=FALSE}
sink("R/data.R")
cat("
#' Mean maximum temperature (monthly average) at Tuggeranong
#'
#' A dataset containing the average mean maximum temperature per month from 1997 to 2019.
#'
#' @format A data frame with 23 rows and 13 variables:
#' \\describe{
#'   \\item{Year}{Year}
#'   \\item{Jan}{January mean maximum temperature}
#'   \\item{Feb}{February mean maximum temperature}
#'   \\item{Mar}{March mean maximum temperature}
#'   \\item{Apr}{April mean maximum temperature}
#'   \\item{May}{May mean maximum temperature}
#'   \\item{Jun}{June mean maximum temperature}
#'   \\item{Jul}{July mean maximum temperature}
#'   \\item{Aug}{August mean maximum temperature}
#'   \\item{Sep}{September mean maximum temperature}
#'   \\item{Oct}{October mean maximum temperature}
#'   \\item{Nov}{November mean maximum temperature}
#'   \\item{Dec}{December mean maximum temperature}
#'   ...
#' }
#' @source \\url{http://www.bom.gov.au/climate/data/}
\"tuggeranong\"
")
sink()

```

Don't use `@export` for a dataset. The export is automatic.

```{r, eval=FALSE}
#roxygenise(".", clean = TRUE)
load_all(".")
document(".")
build(pkg = ".", manual = TRUE)
check(".", manual = TRUE )
install("../temperature", reload = TRUE)

```

```{r}
library(temperature)
data("tuggeranong")
tuggeranong
image(as.matrix(tuggeranong[,-1]))
```

At least two reasons to include data in your package:

1. Give realistic input to examples in your function documentation 
2. Share data along your analyses so that analyses can be reproduced and improved

## Connect to other packages

So far our package only uses R-base code. However we could write fancier functions if we used code already available in other packages.

Let's try to include this function in the package:
```{r}
Wide_To_Long_Temperature <- function(dat, id.vars="Year", 
                                variable.name = "Month",
                                value.name = "Temperature"){
  require(reshape2)
  stopifnot(expr={is.data.frame(dat)
    id.vars %in% colnames(dat)
    })
  melttemp <- melt(dat, id.vars = id.vars,
                 variable.name = variable.name, 
                 value.name = value.name)
  return(melttemp)
}

Wide_To_Long_Temperature(tuggeranong)

```


```{r, eval=FALSE}
#' Wide to long conversion for standard monthly temperature
#'
#' This function takes a standard monthly temperature dataframe in wide format to long format.
#'
#' @param dat A temperature dataframe
#' @param id.vars Character string. vector of id variables. Can be integer (variable position) or string (variable name). If blank, will use all non-measured variables.
#' @param variable.name Character string. Nname of variable used to store measured variable names
#' @param value.name Character string. Name of variable used to store values
#'
#' @return A dataframe
#'
#' @examples
#' data("tuggeranong")
#' Wide_To_Long_Temperature(tuggeranong)
#'
#' @export
Wide_To_Long_Temperature <- function(dat, id.vars="Year",
                                     variable.name = "Month",
                                     value.name = "Temperature"){
  require(reshape2)
  stopifnot(expr={is.data.frame(dat)
    id.vars %in% colnames(dat)
  })
  melttemp <- melt(dat, id.vars = id.vars,
                   variable.name = variable.name,
                   value.name = value.name)
  return(melttemp)
}

```


```{r, eval=FALSE}
load_all()
document()
check()
```
We get a warning message.
Never use library() or require() in a R package!

Two things to do:

1. In DESCRIPTION add:
```{}
Imports:
	reshape2
```
2. In your function change `melt` to `reshape2::melt` 

```{r}
load_all()
document()
check()
```
No more warnings!

Now practice by adding the function below to your package:
```{r}

plot_trend <- function(dat=tuggeranong){
  melttemp <- Wide_To_Long_Temperature(dat)
  
  model <- summary(lm(Temperature ~ Year + Month, data = melttemp))
  year_effect <- round(model$coefficients["Year",],3)
  
  library(ggplot2)
  ggplot(data=melttemp, 
         aes(x=Year,y=Temperature, color=Month))+
    geom_point() + geom_smooth(alpha=0.2) +
    annotate("label", x = mean(melttemp$Year), 
             y = max(melttemp$Temperature), 
             label = paste0(ifelse(year_effect[1]>=0, "+", "-"),
                            year_effect[1], "°C/year, p=",year_effect[4])) 
}

```


##  Make the package a GitHub repository and install from github

If you already have a Github account create a new repo for your package. You can leave the repository completely empty during initialization.

```{}
echo "# testtemperature" >> README.md
git init
git add README.md
git commit -m "first commit"
git remote add origin https://github.com/timotheenivalis/testtemperature.git
git push -u origin master
```

```{}
git add data/ man/ R/ NAMESPACE DESCRIPTION .gitignore .Rbuildignore
git commit -m "load all package files"
git push
```

Now people can install your package from github.
```{r}
install_github("timotheenivalis/testtemperature")
```


## More information

R packages book by Hadley Wickham: http://r-pkgs.had.co.nz/intro.html
